\section{Motivation}
\label{sec:introduction}

Session types enhance the expressivity of traditional types, enabling
to express structured communication on heterogeneously typed
channels~\cite{DBLP:conf/concur/Honda93,DBLP:conf/parle/TakeuchiHK94}.
% Session types have been extended to deal with many realistic
% situations, e.g., multi-party session
% types~\cite{DBLP:conf/popl/HondaYC08}, session types for distributed
% object-oriented programming~\cite{DBLP:conf/popl/GayVRGC10}, session
% types for programming web
% services~\cite{DBLP:journals/toplas/CarboneHY12}.
Thiemann and
Vasconcelos~\cite{thiemann2016context} proposed {\it context-free
  session types} as an extension of session types by allowing nested
protocols that are not restricted to tail recursion. Context-free
session types capture the type-safe serialization of recursive
datatypes and enable the type-safe implementation of remote operations
on recursive datatypes.

Inspired by the context-free session types' framework, Almeida and
Vasconcelos~\cite{bernardo} proposed a functional programming language
equipped with context-free session types.  Such a programming language
is highly dependent on an algorithm to decide type equivalence. We
have developed and implemented an algorithm to decide type
equivalence. Our work capitalizes on the metatheory of context-free
session types proposed by Thiemann and
Vasconcelos~\cite{thiemann2016context}, where type equivalence was
proved to be decidable. Although the decidability of equivalence on
context-free session types has been addressed in the
literature~\cite{DBLP:journals/iandc/ChristensenHS95,janvcar1999techniques,thiemann2016context},
to the best of our knowledge, no algorithm was ever implemented.
% and a possible implementation was not obvious.

The main contributions of this work are: the proposal and implementation of an algorithm to decide
type equivalence of context-free session types, a proof of soundness and completeness of
the algorithm against the declarative definition,
in~\cite{thiemann2016context}, % study and possibly prove the
% completeness of the algorithm
and the validation of  the algorithm on several meaningful examples.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
